A Formal Model of Checked C

### 2023-01-09 sandp23/formal.tex
[review] till function types

This section describes the formal core model of~\systemname,
named~\lang.
We precisely present its syntax, semantics, type
system, and compilation.
The formal model also enables us to develop~\lang’s meta-
theories, including the type soundness, non-exposure, non-crashing,
and the compiler simulation theorems, which we present in~\apdx{sec:theorem}.

@@

\subsection{Syntax}\label{sec:syntax}
The~\fig{fig:checkc-syn} shows a simplified syntax of~\lang.
\subsubsection{Types ($\omega$)}
At a high level, we classify types as word-size values, multi-word values, or functions.
The word-size values can be either integers or pointers.
Pointer types ($\tptr{\omega}{\xi}$) include a pointer mode annotation 
($\xi$, the difference between context and pointer modes is introduced shortly below) that is either checked (\cmode), tainted (\tmode), or unchecked (\umode),
and a type ($\omega$) denoting valid values that can be pointed to.

Multi-word values represent arrays and null-terminated arrays, which include both the type of elements ($\tau$) and a bound ($\bvar$).
The bounds can be either integer literals $n$, expressions $x + n$, or an explicit specification of lower and upper bound,~\ie $(b_l,b_h)$.
We denote null-terminated arrays using $\kappa$, which is $nt$ for null-terminated arrays, and $\cdot$ otherwise (we elide $\cdot$ when writing types).
An example representation of an array and null-terminated array in~\lang syntax is shown below:
\[\hspace*{-0.5em}
\begin{array}{l}
\begin{array}{rcl}
$\code{_t_Array_ptr<}$\tau$\code{> : count(}$n$\code{)}$
&\Leftrightarrow& \tarrayptr{0}{n}{\tau}{\tmode}
\\[0.2em]
$\code{_NT_Array_ptr<}$\tau$\code{> : count(}$n$\code{)}$
&\Leftrightarrow& \tntarrayptr{0}{n}{\tau}{\cmode}
\end{array}
\end{array}
\]
To simplify the representation, we write $\tptr{\tarrayb{b}{\tau}}{\cmode}$ to mean
$\tptr{\tarray{0}{b}{\tau}}{\cmode}$, so the above examples could
be rewritten $\tptr{\tarrayb{n}{\tau}}{\cmode}$ and
$\tptr{\tntarrayb{n}{\tau}}{\cmode}$, respectively.

\emph{Disallowing unsafe types.}
The well-formed requirements of these types are presented in~\fig{fig:wftypesandbounds} (Appendix).
These requirements add additional restrictions for type creation and restrict the creation of unsafe types.
For instance, we cannot create a tainted array whose elements are checked pointers,~\eg Although syntactically valid, the type~\code{_t_Array_ptr<_Ptr<int>>} is~\emph{not} well-formed and is not allowed in~\lang.
However, we can have a checked array whose elements are tainted pointers,~\eg~\code{_t_Array_ptr<_t_Ptr<int>>} is a valid type.

Function types are represented using a dependent function declarations,~\ie $\tfun{\overline{x}}{\overline{\tau}}{\tau}$,
where $\overline{x}$ represents a list of \tint{} type variables that bind bound variables appearing in $\overline{\tau}$ and $\tau$.
An example of a function pointer type is shown below:
\[\hspace*{-0.5em}
\begin{array}{l}
$\code{_t_ptr<(int)(_t_NT_Array_ptr<}$\tau$\code{> : count(}$n$\code{),}$\\
\qquad\qquad$\code{_NT_Array_ptr<}$\tau$\code{>: count(}$n$\code{))>}$
\\[0.2em]
\Leftrightarrow\;\; $\tptr{(\tfun{n}{ \tntarrayptr{0}{n}{\tau}{\tmode} \times \tntarrayptr{0}{n}{\tau}{\tmode}}{\tint})}{\tmode}$
\end{array}
\]
The function type also has well-formed requirements (\fig{fig:wffunctions} in Appendix), which disallows nesting checked or unchecked pointers inside tainted pointers.
Furthermore, these requirements also ensure that all variables in $\overline{\tau}$ and $\tau$ are bounded by $\overline{x}$.
\aravind{The syntax of well-formed requirements for functions is incorrect. It looks like the forall qualifier is missing.}

>>>>

This section describes the formal core model of~\systemname,
named~\lang.
We precisely present its syntax, semantics, type
system, and compilation.
% 
\mzs{The formal model also enables us to}
% 
\mzr{From those, we} develop~\lang’s meta-
theories, including the type soundness, non-exposure, non-crashing,
and the compiler simulation theorems, which we present in~\apdx{sec:theorem}.

@@

\subsection{Syntax}\label{sec:syntax}
The~\fig{fig:checkc-syn} shows a simplified syntax of~\lang.
\subsubsection{Types ($\omega$)}
At a high level, we classify types as word-size value \mzr{type}s, multi-word
value \mzr{type}s, or function \mzr{type}s.
%
\mz{
  % 
  You cannot classify a type as values. You can only classify types into other 3
  categories, namely, word, multi-word and function types.
  % 
}
% 
\mzs{The}\mzr{A} word-size value\mzs{s} can be either \mzr{of type integer} or
\mzr{of type pointer}.
% 
\mzr{Each} pointer type\mzs{s} ($\tptr{\omega}{\xi}$) \mzr{is represented by} a
pointer mode annotation ($\xi$, the difference between context and pointer modes
is introduced shortly below) that is either checked (\cmode), tainted (\tmode),
or unchecked (\umode), and a type ($\omega$) denoting \mzr{the valid type of
  values it can refer to}.
%

% 
\mzr{The multi-word value type ($ \tallarrayb{\bvar}{\tau}$) that ranges over 
arrays and null-terminated arrays is constructed by the type of elements in the
array ($\tau$), an array bound ($\bvar$) and an array flag ($\kappa$).}
% 
\mzr{A} bound can be either \mzr{an} integer literal $n$, \mzr{an} expressions
$x + n$, or an explicit specification of lower and upper bound\mzr{s},~\ie
$(b_l,b_h)$.
%
\mzr{ %
  Additionally, to distinguish null-terminated arrays from other arrays, we set
  \(\kappa\) to be \textit{nt} for nt-arrays and \(\cdot\) for other arrays (for
  convenience, we will elide \(\cdot\) in the text).
  % 
}
%

% 
An example representation of an array and null-terminated array in~\lang syntax is shown below:
% 
\[\hspace*{-0.5em}
\begin{array}{l}
\begin{array}{rcl}
$\code{_t_Array_ptr<}$\tau$\code{> : count(}$n$\code{)}$
&\Leftrightarrow& \tarrayptr{0}{n}{\tau}{\tmode}
\\[0.2em]
$\code{_NT_Array_ptr<}$\tau$\code{> : count(}$n$\code{)}$
&\Leftrightarrow& \tntarrayptr{0}{n}{\tau}{\cmode}
\end{array}
\end{array}
\]
\mzr{For simplicity}\mzs{To simplify the representation}, we write
$\tptr{\tarrayb{b}{\tau}}{\cmode}$ to mean $\tptr{\tarray{0}{b}{\tau}}{\cmode}$,
so the above examples could be rewritten $\tptr{\tarrayb{n}{\tau}}{\cmode}$ and
$\tptr{\tntarrayb{n}{\tau}}{\cmode}$, respectively.

\emph{\mzr{Rejecting}\mzs{Disallowing} unsafe types.}
The well-formed requirements of these types are presented in~\fig{fig:wftypesandbounds} (Appendix).
% 
These requirements \mzs{add additional restrictions for type creation and restrict the creation of unsafe types.}
\mzr{prevents unsafe types from being constructed}.
% 
\mzr{
  % 
  Consider the type ~\code{_t_Array_ptr<_Ptr<int>>} which describes a tainted
  array of checked pointers.
  % 
  Syntactically valid, this type is however \emph{illegal} in~\lang{} and
  rejected by the well-formedness requirement. 
}
% 
\mz{Made some changes to shift the emphasis in hope of improving clearness.}
% 
However, we can have a checked array whose elements are tainted pointers\mzs{,}\mzr{:}~\eg~\code{_t_Array_ptr<_t_Ptr<int>>} is a valid type.
%

% 
Function types are represented using a dependent function declarations,~\ie $\tfun{\overline{x}}{\overline{\tau}}{\tau}$,
where $\overline{x}$ represents a list of \tint{} type variables that bind bound variables appearing in $\overline{\tau}$ and $\tau$.
An example of a function pointer type is shown below:
\[\hspace*{-0.5em}
\begin{array}{l}
$\code{_t_ptr<(int)(_t_NT_Array_ptr<}$\tau$\code{> : count(}$n$\code{),}$\\
\qquad\qquad$\code{_NT_Array_ptr<}$\tau$\code{>: count(}$n$\code{))>}$
\\[0.2em]
\Leftrightarrow\;\; $\tptr{(\tfun{n}{ \tntarrayptr{0}{n}{\tau}{\tmode} \times \tntarrayptr{0}{n}{\tau}{\tmode}}{\tint})}{\tmode}$
\end{array}
\]
The function type also has well-formed requirements (\fig{fig:wffunctions} in Appendix), which disallows nesting checked or unchecked pointers inside tainted pointers.
Furthermore, these requirements also ensure that all variables in $\overline{\tau}$ and $\tau$ are bounded by $\overline{x}$.
\aravind{The syntax of well-formed requirements for functions is incorrect. It looks like the forall qualifier is missing.}
